---
title: "Report on association of systolic blood pressure and death with data from the Framingham study"
author: "Nicole Schönenberger"
date: "`r date()`"
format:
  html:
    self-contained: true
    code-link: true
    code-fold: true
    code-summary: "Code"
    toc: true
    df-print: paged
    embed-resources: true
---

## Introduction

This short report studies the systolic blood pressure in the dataset of the Framingham study and its association with death. The Framingham data can be obtained from the RiskCommunicator package of R. Details of this package can be found in: <https://cran.r-project.org/web/packages/riskCommunicator/index.html>. In this analysis, only the baseline data (timepoint = 0) is used.

## Data import, data preparation, and loading libraries

In a first step, data is imported and only the baseline data is selected. Additionally, libraries are loaded.

```{r}
#| output: false

# Loading packages
library(riskCommunicator)
library(tidyverse)

# Framingham heart study data
framData <- riskCommunicator::framingham

# We keep only the baseline data
framData_base <- subset(framData, TIME == 0)

```

**Data preparation:**

```{r}
#Coding death as data
framData_base <- framData_base |> 
  mutate(DEATH =factor(DEATH, levels=c(0,1),labels=c("not dead","dead")))
```

## Descriptive statistics

In this section, the descriptive statistics of the systolic blood pressure are calculated. This includes sample mean, median, quartiles, sample variance, sample standard deviation, range, and interquartile range (IQR). The descriptive statistics are shown separately for the group that survived and the one that died during the study period.

**Mean of surviving individuals:**

```{r}
# Calculating the mean
mean_bp_d0 <- mean(framData_base$SYSBP[framData_base$DEATH == "not dead"],
                   na.rm = TRUE)
mean_bp_d0
```

**Mean of deceased individuals:**

```{r}
# Calculating the mean
mean_bp_d1 <- mean(framData_base$SYSBP[framData_base$DEATH == "dead"],
                   na.rm = TRUE)
mean_bp_d1
```

**Median of surviving individuals:**

```{r}
# Calculating the median
median_bp_d0 <- median(framData_base$SYSBP[framData_base$DEATH == "not dead"],
                   na.rm = TRUE)
median_bp_d0
```

**Median of deceased individuals:**

```{r}
# Calculating the median
median_bp_d1 <- median(framData_base$SYSBP[framData_base$DEATH == "not dead"],
                   na.rm = TRUE)
median_bp_d1
```

**Quartiles of surviving individuals:**

```{r}
# Calculating quartiles
quart_bp_d0 <- quantile(framData_base$SYSBP[framData_base$DEATH == "not dead"], probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
quart_bp_d0
```

**Quartiles of deceased individuals:**

```{r}
# Calculating quartiles
quart_bp_d1 <- quantile(framData_base$SYSBP[framData_base$DEATH == "dead"], probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
quart_bp_d1
```

**Sample variance of surviving individuals:**

```{r}
# Calculating sample variance
var(framData_base$SYSBP[framData_base$DEATH == "not dead"], na.rm = TRUE)
```

**Sample variance of deceased individuals:**

```{r}
# Calculating sample variance
var(framData_base$SYSBP[framData_base$DEATH == "dead"], na.rm = TRUE)
```

**Standard deviation of surviving individuals:**

```{r}
# Calculating standard deviation
sd_bp_d0 <- sd(framData_base$SYSBP[framData_base$DEATH == "not dead"], na.rm = TRUE)
sd_bp_d0 
```

**Standard deviation of deceased individuals:**

```{r}
# Calculating standard deviation
sd_bp_d1 <- sd(framData_base$SYSBP[framData_base$DEATH == "dead"], na.rm = TRUE)
sd_bp_d1 
```

**Range of systolic blood pressure of surviving individuals:**

```{r}
# Range of systolic blood pressure
range(framData_base$SYSBP[framData_base$DEATH == "not dead"], na.rm = TRUE)
```

**Range of systolic blood pressure of deceased individuals:**

```{r}
# Range of systolic blood pressure
range(framData_base$SYSBP[framData_base$DEATH == "dead"], na.rm = TRUE)
```

**Interquartile range of surviving individuals:**

```{r}
# Calculating interquartile range
IQR(framData_base$SYSBP[framData_base$DEATH == "not dead"], na.rm = TRUE)
```

**Interquartile range of deceased individuals:**

```{r}
# Calculating interquartile range
IQR(framData_base$SYSBP[framData_base$DEATH == "dead"], na.rm = TRUE)
```

**Summary of descriptive statistics of surviving individuals:**

```{r}
# Summary of descriptive statistics
summary(framData_base$SYSBP[framData_base$DEATH == "not dead"])
```

**Summary of descriptive statistics of deceased individuals:**

```{r}
# Summary of descriptive statistics
summary(framData_base$SYSBP[framData_base$DEATH == "dead"])
```

## Visual representation of systolic blood pressure

[Fig-1](#fig-1) illustrates the systolic blood pressure in the framingham study population. The figure shows evidence that the systolic blood pressure is not normally distributed and that the data is skewed in both groups. When performing the Shapiro test in both groups the p-value is 2.2e-16, meaning that the data is not normally distributed. This is also visualized by the QQplots shown in [Fig-2](#fig-2) because the sample quantiles differ from the theoretical quantiles. In the boxplots [Fig-3](#fig-3) it is visualized that the systolic blood pressure in the deceased individuals were higher then the one in surviving individuals.

```{r}
#| label: Fig-1
#| fig-cap: "Histograms of the systolic blood pressure of the group that survived the study duration (left histogram) and the individuals that died (right histogram). The solid lines represent the medians, whereas the dashed lines represent the respective means."
#| fig-width: 8
#| fig-height: 4
# Plot distribution of blood pressure separately for patients that died and patients that did not
ggplot(framData_base, aes(x = SYSBP, color = DEATH)) + 
  geom_histogram(aes(y = after_stat(density)), breaks = seq(0, 250, by = 5), fill = "grey") +
  geom_vline(aes(xintercept = median(SYSBP)), color = "black", linetype = "solid") +
  geom_vline(aes(xintercept = mean(SYSBP)), color = "black", linetype = "dashed") +
  facet_wrap(~DEATH) +
  scale_color_manual(values = c("#999999", "#E69F00"))

```

**Shapiro test for surviving individuals:**
```{r}
# Shapiro test for surviving individuals
shapiro_survived <- shapiro.test(framData_base$SYSBP[framData_base$DEATH == "not dead"])
shapiro_survived
```

**Shapiro test for deceased individuals:**
```{r}
# Shapiro test for deceased individuals
shapiro_deceased <- shapiro.test(framData_base$SYSBP[framData_base$DEATH == "dead"])
shapiro_deceased

```
```{r}
#| label: Fig-2
#| fig-cap: "QQplots of the systolic blood pressure of both groups (surviving and deceased)."
#| fig-width: 8
#| fig-height: 4
# QQplot for surviving and deceased individuals
ggplot(framData_base, aes(sample = SYSBP)) +
  geom_qq(color = "black") +
  geom_qq_line(color = "red") +
  facet_wrap(~ DEATH) +
  labs(title = "QQplots of Systolic Blood Pressure",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()

# Boxplot with ggplot2
#| label: Fig-3
#| fig-cap: "Box plots of the systolic blood pressure of both groups (surviving and deceased)."
#| fig-width: 8
#| fig-height: 4
ggplot(framData_base, aes(y = SYSBP, groups = DEATH, color = DEATH)) + 
  geom_boxplot() +
  scale_x_discrete() +
  scale_color_manual(values = c("#999999", "#E69F00"))
```

## Wilcoxon test
To analyze if the observed difference in median systolic blood pressure values are statistically significant, the Wilcoxon test was performed. The result of the Wilcoxon test suggests that there is evidence to support the presence of a significant difference in systolic blood pressure between deceased and surviving individuals in the studied dataset (difference in location = -12.5, 95% CI = -14.00 to -11.00, p=<2.2e-16). 

```{r}
#| output: false
# Wilcoxon test
data_dead <- subset(framData_base, DEATH == "dead")
data_not_dead <- subset(framData_base, DEATH == "not dead")

wilcox_result <- wilcox.test(data_dead$SYSBP, data_not_dead$SYSBP, alternative = "two.sided", correct = TRUE)
# Print results
cat("Wilcoxon Test Result:")
print(wilcox_result)
```
Wilcoxon rank sum test with continuity correction

data:  framData_baseSYSBP by framData_baseDEATH
W = 1471742, p-value < 0.00000000000000022
alternative hypothesis: true location shift is not equal to 0
95 percent confidence interval:
 -13.99993 -11.00005
sample estimates:
difference in location 
                 -12.5 
                 
## Conclusion
The descriptive statistics of systolic blood pressure for the group that survived and the group that died during the study period have been calculated. These statistics provide an overview of the central tendency, dispersion, and range of systolic blood pressure in each group.There is evidence that the systolic blood pressure in the Framingham study population was higher in deceased individuals versus the individuals that survived. The Wilcoxon test was chosen for the analysis because the systolic blood pressure was not normally distributed. However, as there are 4434 individuals, a t-test would also have been appropriate to use because of this large sample size.

